Unicode Implementation Issues \thanks { This is a working document for the Rexx Language Association (RexxLA). There will be one or more separate requirements documents. It is written in \LaTeX\ \fmtversion\ and was rendered \today\ using \enginedetails. }

### 2023-09-26 ARB/standards/work-in-progress/unicode/Unicode.Implimentation.tex
Move implementation document to its proper place

% \documentclass[draft]{article}
\documentclass[b4paper]{article}
\usepackage{enumitem}
\usepackage{eqparbox}
\usepackage{expl3}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{listings}         %Requires luatex engine to handle UTF-8
                              %Consider using minted, but ...
\usepackage{upquote}

\newlist{definition}{description}{2} %% <- pick a larger number if you want to nest these
\setlist[definition]{
    labelwidth=\eqboxwidth{listlabel@\EnumitemId},
    leftmargin=!,
    format=\definitionlabel,
}
\newcommand\definitionlabel[2][l]{\eqmakebox[listlabel@\EnumitemId][#1]{#2}}

\newlist{resource}{description}{2} %% <- pick a larger number if you want to nest these
\setlist[resource]{
    labelwidth=\eqboxwidth{listlabel@\EnumitemId},
    leftmargin=!,
    format=\resourcelabel,
}
\newcommand\resourcelabel[2][l]{\eqmakebox[listlabel@\EnumitemId][#1]{#2}}

@@

\section{Background}
Many modern languages support Unicode encoded as
\href{https://datatracker.ietf.org/doc/rfc3629/}{UTF-8} (\cite{RFC 3629}),
and the \href{https://www.ietf.org/}{IETF}, in \cite{RFC 5198},
has mandated the use of UTF-8 with NFC in new protocols.
Rexx will need to support UTF-8 if it is not to become a backwater,
and that support should not break existing code.

\section{Scope}
This document only describes issues related to expanded alphabets and multi-octet representations of characters;
it does not address such issues as bidirectional text.
It presents suggested solutions for requirements that are presented in other documents.
Also, most of the details are under active discussion, so everything here is provisional.

@@

The definitions given in \cite{Glossary} and \cite{Unicode} take
precedence over those given here. Quoted text is taken from those sources.
Except for definitions taken from official IETF and Unicode documents,
the nomenclature used here is illustrative rather than normative;
language design teams will formally define, e.g., method names, encoding of parameters.
Most of the sections assume that there will be destinct string types
for extended grapheme clusters, Unicode code points, legacy code pages
and raw octets. However, there has been discussion of including
metadata, in which case come of the classes might be merged.

\begin{definition}
\item [.Legacy] The string class for strings in legacy code pages.
\item [.octets] The string class for uninterprted octet strings.
\item [.Ucp] The string class for strings of Unicode code points.
\item [.Ugc] The string class for strings of Unicod extended grapheme clusters.
\item [\textbf{\'{ }}] U+0301, COMBINING ACUTE ACCENT
\\*
Set in bold to distinguish it from U+27, APOSTROPHE.
\item [Bidi] "Abbreviation of bidirectional, in reference to mixed left-to-right and right-to-left text."
\item [BIF] Built In Function
\item [BIM] Built In Method
\item [Block] "A grouping of characters within the Unicode encoding
space used for organizing code charts. Each block is a uniquely named,
continuous, non-overlapping range of code points, containing a
multiple of 16 code points, and starting at a location that is a
multiple of 16. A block may contain unassigned code points, which are
reserved."
\item [BMP] Basic Multilingual Plane
\\*
The first 64 Ki code points of Unicode, from U+0000 to U+FFFF.
\item [BOM] Byte Order Mark: U+FEFF, ZERO WIDTH NON-BREAKING SPACE (ZWNBSP)
\\*
Used as the first character to indicate byte order for UCS-2, UTF-16 and UCS-4;
\\*
Optional as the first character for UTF-8.
\item [\'{E}] U+C9, LATIN CAPITAL LETTER E WITH ACUTE
\\*
Set in normal weight: U+45 U+0301 will be shown as "E\textbf{\'{ }}".
\item [\'{e}] U+E9, LATIN SMALL LETTER E WITH ACUTE
\\*
Set in normal weight: U+65 U+0301 will be shown as "e\textbf{\'{ }}".
\item [encoded character] The smallest constituant of a Unicode string.
"The Unicode Standard does not define what is and is not a text
element in different processes; instead, it defines elements called
encoded characters. An encoded character is represented by a number
from 0 to 10FFFF\textsubscript{16} called a code point."
\item [EGC] Extended grapheme cluster
\item [GCGID] "Acronym for Graphic Character Global Identifier.
These are listed in the IBM document
Character Data Representation Architecture, Level 1, Registry SC09-1391."
See \url{https://www.ibm.com/downloads/cas/G01BQVRV}.
\item [grapheme cluster]
"A grapheme cluster consists of a base character followed by any
number of continuing characters, where acontinuing character may
include any nonspacing combining mark, certain spacing combining
marks, or a join control."
{\cite{Annex29}} defines two types of grapheme clusters;
"An extended grapheme cluster is the same as a legacy grapheme cluster,
with the addition of some other characters.
The continuing characters are extended to include all spacing combining marks,
such as the spacing (but dependent) vowel signs in Indic scripts."
\item [high surrogate] A code point in the range U+D800-U+DBFF, used
as the first half of a surrogate pair.
\item [IETF] \href{https://ietf.org}{Internet Engineering Task Force}
\item [introducer] The first octet in the UTF-8 encoding of a Unicode character beyond U+7F
\item [low surrogate] A code point in the range U+DC00-U+DFFF, used
as the second half of a surrogate pair.
\item [NFC]  Unicode
"Normalization Form C (NFC). A normalization form that erases any
canonical differences, and generally produces a composed result. For
example, a + umlaut is converted to \symbol{"00E4} in this form. This form most
closely matches legacy usage. The formal definition is D120 in Section
3.11, Normalization Forms."
\\*
the normalization endorsed by the IETF
\item [NFD] Unicode
"Normalization Form D (NFD). A normalization form that erases any
canonical differences, and produces a decomposed result. For example,
\symbol{"00E4} is converted to a + umlaut in this form. This form is most often
used in internal processing, such as in collation. The formal
definition is D118 in Section 3.11, Normalization Forms."

\item [octet] 8-bit byte
\item [Plane] "A range of 65,536 (1000016) contiguous Unicode code
points, where the first code point is an integer multipleof 65,536
(1000016). Planes are numbered from 0 to 16, with the number being the
first code point of the plane divided by 65,536. Thus Plane 0 is
U+0000..U+FFFF, Plane 1 is U+10000..U+1FFFF, ..., and Plane 16 (1016)
is U+100000..10FFFF. (Note that ISO/IEC 10646 uses hexadecimal
notation for the plane numbers-for example, Plane B instead of Plane
11). (See Basic Multilingual Plane and supplementary planes.)"
\item [RFC] Request For Comments
\\*
A formal document published by the IETF defining, e.g., a protocol.
RFC documents contain technical specifications and organizational notes for the Internet.
\begin{itemize}
  \item Best Current Practice (BCP)
  \item Experimental
  \item Informational
  \item Proposed Standard
  \item Internet Standard (STD)
  \item Historic
\end{itemize}
\item [surrogate] A code point in the range U+D800-U+DFFF used to
encode 21-bit code points into pairs of 16-bit bytes.
\item [surrogate pair] A high surrogate (in the range U+D800-U+DBFF)
followed by a low surrogate (in the range U+DC00-U+DFFF),
collectively representing a 21-bit code point.
\item [TBD] To Be Determined.
\item [UCS] Universal Character Set, ISO 10646, roughly equivalent to Unicode
\item [UCS-2] A 16 bit subset of Unicode, containing only the BMP.
\item [The Unicode Consortium] A non-profit corporation devoted to
developing, maintaining, and promoting software internationalization standards and data.
\item [UTF-8] UCS Transformation Format 8 (\cite{RFC 3629}).
\\*
The encoding of Unicode endorsed by the IETF
\item [UTF-8 octet sequence]
The sequence of octets representing a single Unicode code point.
It may consist of a single ASCII character padded on the left with a zero bit,
or of a one octet introducers followed by a 1-3 octet tail.

\begin{tabular}{| r | l |}
\hline
Code points & UTF-8 octet sequence \\
\hline
U+0000 - U+007F    & 0xxxxxxx \\
\hline
U+0080 - U+07FF    & 110xxxxx 10xxxxxx \\
\hline
U+0800 - U+FFFF    & 1110xxxx 10xxxxxx 10xxxxxx \\
\hline
U+10000 - U+10FFFF & 11110xxx 10xxxxxx 10xxxxxx 10xxxxxx \\
\hline
\end{tabular}
\item [UTF-16] UCS Transformation Format 16 (\cite{RFC 2781}).
\end{definition}

@@

\section{Statement of problem}

\subsection{Primary Problem}

In Rexx, everything is a string; there are no classes, declarations or types.

While the ANSI standard does not mandate any particular character width for \textbf{Config\_C2B()},
most if not all implementations use a width of 8,
and the length of \textbf{c2x(}\textit{foo}\textbf{)}
is twice the length of \textit{foo}.

A large body of existing code operates on binary data from external files,
or accessed from memory via the \textbf{storage()} BIF,
under the assumption that, e.g., \textbf{c2x()}, left(), \textbf{right()}, \textbf{substr()}, operate on octets.
A UTF-8 introducer is treated the same as any other value.

A large body of existing code operates on text under the assumption that those facilities operate on characters.

@@

For example, the Unicode string 'Caf\'{e}' has 4 code points, 4
grapheme clusters and 5 octets in UTF-8 encoding while
'Cafe\textbf{\'{ }}' has 5 code points, 4 grapheme clusters and 6 octets
in UTF-8 encoding, yet many text processing applications need to treat
them as equivalent.

\subsection{Secondary Problems}
Rexx has built-in case conversion, but it is based on ASCII and
can't even convert the accented letters found in, e.g., ISO-8859-1
(Latin-1), ISO-8859-15 (Latin-9), much less all those found in Unicode.

@@

In current implementations the range of characters is extremely small,
and thus the \textbf{xrange()} only returns short strings.
With Unicode the range expands to U+00 through U+10FFFF.
That is far too large to allow as either an explict or implicit range in \textbf{xrange()}.

The current definition restricts symbols to ASCII characters,
which excludes letters used in many languages. If the rules are extended to include
Unicode letters and digits beyond ASCII, the rules for equality of symbols must be addressed.
Are composed and decomposed strings identical?
Are base characters identical to their compatibility alterrnates?
Are subscript and superscript digits distinct from their ASCII
counterparts?

When dealing with Unicode, different code sequences may have identical
rendering due to the existence of both fully composed and combining
code points and the exixtence of compatibility code points.
Rexx needs a way to test two Unicode strings for equivalence.

There has been some discussion claiming a need to continue supporting legacy code pages.

There has been some discussion of the need to detect unassigned code points.

@@

Most of this section assumes that there will be three distict string
types, and that some methods will not exist in all three types, or
will behave differently.
The type names used here are placeholders, and will be replaced once
there is consensus on what to call them.

@@

OPTIONS mighy also specify a source encoding parameter, overriding any code page in an environment variable or file metadata.

\subsection{Types}

Allow Rexx variables to contain three distinct types of data: raw octets, Unicode code points and text,
Define \textbf{storage()} as returning raw octets,
and provide conversion functions ("casts").
There is an ongoing discussion as to whether all three are needed.

There should be no implicit conversion between \textbf{.octets} and Unicode strings.

There should be implicit conversion between Unicode types, but it need not be reversible, i.e.,
a conversion from .Ucp to .Ugc and back need not produce the same code points as the original

@@

\subsection{Width parameter}

Add a width parameter to \textbf{c2x()} for Unicode code point string
(raw octet strings may require input and output widths for UCS-2 AND UTF-16 data, and the
utility of \textbf{c2x()} for grapheme clusters needs more
analysis), and raise conditions with distinct error codes if any code
point is out of range or if an invalid UTF-8 sequence is detected.  A
case could be made for using either the bit size or the digit size as
the width.

@@

There is an ongoing discussion as to whether an when to allow implicit coercions of type in, e.g.,
\begin{lstlisting}[linewidth=35em]
    foo = .octets~new
    foo = 'Ren!\'{e}!'
    !\textellipsis!
    parse var foo ASCII !{\textquotesingle}\'{e}{\textquotesingle}! .
    say ASCII                     /* Displays Ren */
\end{lstlisting}

@@

\subsection{Constants}
Other languages allow specifying Unicode characters using either
the hexadecimal value of the code point or the assigned name of the code point,
e.g., U+E9 might be coded as {\textbackslash}u\{E9\} or {\textbackslash}u[LATIN SMALL LETTER E WITH ACUTE].
The syntax used for such constants should be consisten across Rexx variants,
including rules for optional spaces between (hex) digits,
and should take into account the recommendations in \cite{RFC 5137}.
Implementations should use the machine readable data bases published by
the Unicode Consortium in order to ease migration to new versions of Unicode.

@@

Binary and hexadecimal literals are of type \textbf{.octets} and cannot be implicitly coerced to Unicode strings.
Other string literals are legacy or Unicode and cannot be implictly coerced to \textbf{.octets}.

@@

\begin{tabular}{| l | l | l |}
\hline
Example & Type & Semantics \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}              & Ucp & Unicode text by code point \\
                                                          & Ugc & Unicode text by grapheme cluster (TBD) \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}R               & Ucp & Unicode text by code point \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}Ucp             & Ucp & Unicode text by code point \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}Ugc             & Ucp & Unicode text by grapheme cluster  \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{ucp\}       & Ucp & Unicode text by code point \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{ugc\}       & Ucp & Unicode text by grapheme cluster  \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{ISO8859-1\} & Legacy & ISO 8859-1 legacy text \\
\hline
\end{tabular}

\subsection{Error detection}
The condition names given below are illustrative and not normative.
Unicode BIF/BIMs should detect invalid input and signal the following conditions, with unique error codes.

@@

The \textbf{command} method of the stream classes should support a \textbf{CODEPAGE} option;
attempting to set a codepage for a binary file should raise a \textbf{NOTTEXT} condition with a distinct error code.
There should be an option or suboption controling whether to use Unicode if an apparent BOM is detected.

@@

@@

The rules for case folding should take into account mathematical usage.
\begin{itemize}
\item U+1D400 through U+1D7FF are semantically distinct from other letters.
\item Superscripts and subscripts have semantic significance.
\end{itemize}

>>>>

%Compile with latexmk -lualatex
%
% Changes
% Table of contents
% \underline{default}
% Split nomenclature section
% Link to 3.9 Unicode Encoding Forms
% Codepoints/EGCs option in support of charin()
% Correct and expand constants example
% Index
% stream - seek and update
% TUTOR
% Reference ARB GitHub issues
%
% \documentclass[draft]{article}
\documentclass[b4paper]{article}
\usepackage{enumitem}
\usepackage{eqparbox}
\usepackage{expl3}
\usepackage{fontspec}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{listings}         %Requires luatex engine to handle UTF-8
                              %Consider using minted, but ...
\usepackage{makeidx}
\usepackage{upquote}
\usepackage{cleveref}

\makeindex
\setmainfont{CMU Serif}
% texlive-cm-unicode or retrieve cm-unicode with MikTex Console

\newlist{definition}{description}{2} % <- pick a larger number if you want to nest these
\setlist[definition]{
    labelwidth=\eqboxwidth{listlabel@\EnumitemId},
    leftmargin=!,
    format=\definitionlabel,
}
\newcommand\definitionlabel[2][l]{\eqmakebox[listlabel@\EnumitemId][#1]{#2}}

\newlist{issue}{enumerate}{2} % <- pick a larger number if you want to nest these
\crefname{issuei}{issue}{issues}
\Crefname{issuei}{Issue}{Issues}
\crefname{issueii}{issue}{issues}
\Crefname{issueii}{Issue}{Issues}

\newlist{resource}{description}{2} % <- pick a larger number if you want to nest these
\setlist[resource]{
    labelwidth=\eqboxwidth{listlabel@\EnumitemId},
    leftmargin=!,
    format=\resourcelabel,
}
\newcommand\resourcelabel[2][l]{\eqmakebox[listlabel@\EnumitemId][#1]{#2}}

@@

\tableofcontents

\section{Background}
The Architecture Review board is investigating areas in which Rexx may be
in need of enhancements, some of which are listed in \cite{Req}.
Many modern languages support Unicode encoded as
\href{https://datatracker.ietf.org/doc/rfc3629/}{UTF-8} (\cite{RFC 3629}),
and the \href{https://www.ietf.org/}{IETF}, in \cite{RFC 5198},
has mandated the use of UTF-8 with NFC in new protocols.
Rexx will need to support UTF-8 if it is not to become a backwater,
and that support should not break existing code.
Elaboration of this issue may be added to \cite{Req} \cref{issue:MB,issue:UTF-8}.

\section{Scope}
This document only describes issues related to \cref{issue:EA,issue:MB}, expanded alphabet and
multi-octet representations of characters in Unicode \cref{issue:MB,issue:UTF-8}; it does not
address such issues as bidirectional text, nor does it address legacy
DBCS support. It presents suggested solutions for requirements that
are presented in other documents.  Also, most of the details are under
active discussion, so everything here is provisional.

This document concentrates on classic Rexx, ANSI Rexx and ooRexx; the
implementations most relevant are ooRexx and Regina.  However, cREXX
and NetRexx must be taken into account in order to avoid unnecessary
differences.

@@

The definitions given in \cite{Glossary} and \cite{Unicode} take
precedence over those given here. Quoted text is taken from those sources.
Except for definitions taken from official IETF and Unicode documents,
the nomenclature used here is illustrative rather than normative;
language design teams will formally define, e.g., method names, encoding of parameters.
Most of the sections assume that there will be destinct string types
for extended grapheme clusters (\cref{issue:NT,issue:c2u}), Unicode code points (\cref{issue:NT,issue:c2u}),
legacy code pages (\cref{issue:Leg,issue:mLS,issue:iconv})
and raw octets (\cref{issue:ROS,issue:NT,issue:mO}). However, there has been discussion of including
metadata, in which case come of the classes might be merged.

Some of the nomenclature used in this document, including
the illustrative class names \textbf{BYTES}, \textbf{CODEPOINTS} and \textbf{TEXT}
is taken from the RXU preprocessor of The Unicode Tools of Rexx (TUTOR).
It is planned to encorporate more vocabulary from RXU in subsequent revisions.

The word \textbf{text} has its generic meaning unless given in upper case.

An \underline{underscored} term in a list of options is the default.
Terms separated by the Or symbol (|) are mutually exclussive alternatives.
Terms in brackets are optional.

The bold character \textbf{\'{ }} in examples refers to
U+0301, COMBINING ACUTE ACCENT.
The characters \textbf{\'{E}} and \textbf{\'{e}} in examples refer to the
composed characters U+C9, LATIN CAPITAL LETTER E WITH ACUTE and
U+E9, LATIN SMALL LETTER E WITH ACUTE, not to the decomposed clusters
<U+0045, U+0301> and <U+0065, U+0301>.

@@

References to requirments in \cref{ARBiss} \nameref{ARBiss} are by number.

\section{Statement of problems}

\subsection{Character width}

In Rexx, everything is a string; there are no classes, declarations or types. (\cref{issue:Leg})

While the ANSI standard does not mandate any particular character width for \textbf{Config\_C2B()},
most if not all implementations use a width of 8,
and the length of \textbf{c2x(}\textit{foo}\textbf{)}
is twice the length of \textit{foo}. (\cref{issue:c2x,issue:c2xW})

A large body of existing code operates on binary data from external files,
or accessed from memory via the \textbf{storage()} BIF,
under the assumption that, e.g., \textbf{c2x()}, left(), \textbf{right()}, \textbf{substr()}, operate on octets(]cref{issue:ROS,issue:mO}.
A UTF-8 introducer is treated the same as any other value.

A large body of existing code operates on text under the assumption
that those facilities operate on characters. (\cref{issue:US})

@@

For example, the Unicode string "Caf\'{e}" has 4 code points, 4
grapheme clusters and 5 octets in UTF-8 encoding while
"Cafe\textbf{\'{ }}" has 5 code points, 4 grapheme clusters and 6 octets
in UTF-8 encoding (\cref{issue:NT}), yet many text processing applications need to treat
them as equivalent.

\subsection{Case conversion}
Rexx has built-in case conversion, but it is based on ASCII and
can't even convert the accented letters found in, e.g., ISO-8859-1
(Latin-1), ISO-8859-15 (Latin-9), much less all those found in Unicode.

@@

\subsection{\textbf{xrange()}}
In current implementations the range of characters is extremely small,
and thus the \textbf{xrange()} only returns short strings. (\cref{issue:xr})
With Unicode the range expands to U+00 through U+10FFFF.
That is far too large to allow as either an explict or implicit range in \textbf{xrange()}.

\subsection{Identifiers, labels and procedure names}
The current definition restricts symbols to ASCII characters, which
excludes letters used in many languages. If the rules are extended, in
accordance with \cite{Annex31}.  to include Unicode letters and digits
beyond ASCII, the rules for equality of symbols must be addressed.
Are composed and decomposed strings identical? Are base characters
identical to their compatibility alterrnates? Are subscript and
superscript digits distinct from their ASCII counterparts?

When a numeric value is used as a label or procedure name, a precise
definition of the semantics is needed.  While the definition
"Translate Binary to a coded string." of Config\_B2C in ANSI X3J18-199X
is rather vague, it is adequate when the character set is restricted
to an 8-bit character set.  The definition must be replaced by one
that is unamgiguous when applied to Unicode.  (\ref{issue:IAF})
There are several
obvious options.
\begin{enumerate}
\item Treat every 32 bits as a Unicode scalar
\item Treat every 8 bits as part of a UTF-8 sequence
\item Treat every 8 bits as a character in a legacy code page.
\end{enumerate}
Each of these options has disadvantages.
The same issue exists for Config\_C2B.

When dealing with Unicode, different code sequences may have identical
rendering due to the existence of both fully composed and combining
code points and the exixtence of compatibility code points.
Rexx needs a way to test two Unicode strings for equivalence. (\cref{issue:SNO,issue:NFCD,issue:NFKCD})

\subsection{Status of legacy code pages}
There has been some discussion claiming a need to continue supporting legacy code pages. (\cref{issue:LCPS,issue:mLS,issue:iconv})

\subsection{Unassigned and invalid code points}
There has been some discussion of the need to detect unassigned code points. (\cref{issue:UICP,issue:ED})

\subsection{New notation}
Unicode introduces new notation, which Rexx currently doesn't support, e.g., U+xxxx, <codepoint, ...>.

\section{ARB Issues}
\label{ARBiss}

This section lists all issues in \cite{Req}, whether or not they relate to Unicode.

\begin{issue}[label=I\arabic*,ref=\fbox{I\arabic*}]
\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730857}%
{Expanded Alphabets}}
\label{issue:EA}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730866}%
{Multi-Byte Representations}}
\label{issue:MB}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730964}%
{Unicode Strings}}
\label{issue:US}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730903}%
{Unassigned \& Invalid Code Points}}
\label{issue:UICP}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730928}%
{Error Detection}}
\label{issue:ED}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730846}%
{UTF-8 Support}}
\label{issue:UTF-8}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730880}%
{Identifiers \& Allowable Formats}}
\label{issue:IAF}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730910}%
{New Notation (e.g., U+xxxx)}}
\label{issue:NN}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730970}%
{Constants for Unicode Characters}}
\label{issue:CUC}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730937}%
{Case Folding}}
\label{issue:CF}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731043}%
{Caseless Comparisons}}
\label{issue:CComp}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730872}%
{Case Conversion}}
\label{issue:CConv}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730976}%
{Coercions}}
\label{issue:C}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731061}%
{Strict / Non-strict operators}}
\label{issue:SNO}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730954}%
{Legacy Strings}}
\label{issue:Leg}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730887}%
{Legacy Code Pages Support}}
\label{issue:LCPS}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730947}%
{Raw Octet Strings}}
\label{issue:ROS}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730934}%
{I/O Methods}}
\label{issue:I/O}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35753880}%
{IO Random Access}}
\label{issue:RA}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35801621}%
{New Text Functions}}
\label{issue:NT}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731024}%
{NFC \& NFD Functions}}
\label{issue:NFCD}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731034}%
{NFKC \& NFKD Functions}}
\label{issue:NFKCD}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730987}%
{c2x() Function}}
\label{issue:c2x}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731048}%
{Width Parameters for c2x()}}
\label{issue:c2xW}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730995}%
{c2u() Function}}
\label{issue:c2u}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731000}%
{makeCodePointString Function}}
\label{issue:mCPS}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731003}%
{makeGraphemClusterString Function}}
\label{issue:mGCS}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731012}%
{makeLegacyString Function}}
\label{issue:mLS}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731020}%
{makeOctets Function}}
\label{issue:mO}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731039}%
{iconv Function}}
\label{issue:iconv}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35730875}%
{xrange() Function}}
\label{issue:xr}

\item{\href{https://github.com/users/RexxLA/projects/2/views/6?pane=issue&itemId=35731055}%
{PARSE Instruction}}
\label{issue:P}

\end{issue}

@@

Most of this section assumes that there will be three distict string
types, (\cref{issue:mCPS,issue:mGCS,issue:mO}),
and that some methods will not exist in all three types, or
will behave differently.
The type names used here are placeholders, and will be replaced once
there is consensus on what to call them.

@@

OPTIONS might also specify a source encoding parameter, overriding any
code page in an environment variable or file metadata.

\begin{definition}
\item[SOURCECP=] Encoding of the Rexx source code
\item[LITERALS=] Type of string constant with no suffix
\begin{definition}
\item[BYTES] Raw octets
\item[CODEPOINTS] Unicode by code point
\item[LEGACY] Legacy 8-bit characters
\item[\underline{TEXT}] Unicode by extended grapheme cluster
\end{definition}
\end{definition}

\subsection{Classes and types}

Allow Rexx variables to contain three distinct types of data: raw octets, Unicode code points and text,
(\cref{issue:mCPS,issue:mGCS,issue:mO}),

Define \textbf{storage()} as returning raw octets,
and provide conversion functions ("casts").
There is an ongoing discussion as to whether all three are needed.

There should be no implicit conversion between \textbf{.BYTES} and Unicode strings.

There should be implicit conversion between Unicode types, but it need not be reversible, i.e.,
a conversion from .CODEPOINTS to .TEXT and back need not produce the same code points as the original

@@

For ooRexx a single class with distinguishing attributes could be
used, or a separate class for each type.

\subsection{Width parameters for c2x()}

Add a width parameter to \textbf{c2x()} for Unicode code point string
(\cref{issue:c2xW})
(raw octet strings may require input and output widths for UCS-2 AND UTF-16 data, and the
utility of \textbf{c2x()} for grapheme clusters needs more
analysis), and raise conditions with distinct error codes if any code
point is out of range or if an invalid UTF-8 sequence is detected.  A
case could be made for using either the bit size or the digit size as
the width.

@@

There is an ongoing discussion as to whether an when to allow implicit coercions of type in, e.g.,
\begin{lstlisting}[linewidth=35em]
    foo = .BYTES~new
    foo = 'Ren!\'{e}!'
    !\textellipsis!
    parse var foo ASCII !{\textquotesingle}\'{e}{\textquotesingle}! .
    say ASCII                     /* Displays Ren */
\end{lstlisting}

@@

\subsection{Constants}
Other languages allow specifying Unicode characters using either
the hexadecimal value of the code point or the assigned name of the code point,
(\cref{issue:CUC})
e.g., U+E9
(\cref{issue:NN})
might be coded as {\textbackslash}u\{E9\} or {\textbackslash}u[LATIN SMALL LETTER E WITH ACUTE].
The syntax used for such constants should be consisten across Rexx variants,
including rules for optional spaces between (hex) digits,
and should take into account the recommendations in \cite{RFC 5137}.
Implementations should use the machine readable data bases published by
the Unicode Consortium in order to ease migration to new versions of Unicode.

@@

Binary and hexadecimal literals are of type \textbf{.BYTES} and cannot be implicitly coerced to Unicode strings.
Other string literals are legacy or Unicode and cannot be implictly coerced to \textbf{.BYTES}.

@@

\begin{tabular}{| l | l | l |}
\hline
Example & Type & Semantics \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}               & CODEPOINTS & Unicode text by code point             \\
                                                           & TEXT       & Unicode text by grapheme cluster (TBD) \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}R                & CODEPOINTS & Unicode text by code point             \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}CODEPOINTS       & CODEPOINTS & Unicode text by code point             \\
\hline
{\textquotesingle}U+00E9{\textquotesingle}TEXT             & TEXT       & Unicode text by grapheme cluster       \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{CODEPOINTS\} & CODEPOINTS & Unicode text by code point             \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{TEXT\}       & TEXT       & Unicode text by grapheme cluster       \\
\hline
{\textquotesingle}Ren\'{e}{\textquotesingle}\{ISO8859-1\}  & Legacy     & ISO 8859-1 legacy text                 \\
\hline                                                     & CODEPOINTS & Unicode text by code point             \\
{\verb|`|}Ren\'{e}{\verb|`|}
                                                           & TEXT       & Unicode text by grapheme cluster (TBD) \\
\hline
{\verb|`|}Ren\'{e}{\verb|`|}C                              & CODEPOINTS & Unicode text by code point             \\
\hline
{\verb|`|}Ren\'{e}{\verb|`|}T                              & TEXT       & Unicode text by grapheme cluster       \\
\hline
\end{tabular}

An issue that must be resolved is whether string literals should have
binary fidelity or visual fidelity to the target code page, i.e.,
should the quoted source characters be given in the source code page
or in the target code page.  E.g., should the word "foo" in ISO 8859-1
and in EBCDIC be coded as
{\textquotesingle}foo{\textquotesingle}{ISO8859-1} and
{\textquotesingle}foo{\textquotesingle}{EBCDIC}, requiring that Rexx
translate the text, or should the second be entered with the EBCDIC
code points 86, 96, 96, making it harder to read and edit?

\subsection{Error detection}
The Rexx standard should specify ehat to do when a string being transformed
contains invalid code points or octet seqiences. (\cref{issue:ED})
The condition names given below are illustrative and not normative.
Unicode BIF/BIMs should detect invalid input and signal the following conditions, with unique error codes.
It is TBD whether some or all of these should be folded into SYNTAX.

@@

Stream input/output will need additional support for Unicode.
\cref{issue:I/O}
The \textbf{command}  and \textbf{open} methods of the stream classes
should support a \textbf{CODEPAGE} option; attempting to set a
codepage for a binary file should raise a \textbf{NOTTEXT} condition
with a distinct error code.  There should be an option or suboption
controling whether to use Unicode if an apparent BOM is detected.

@@

Similarly, there should be an option to distinguish between reading
individual code points and reading EGC; this is needed so that the
\textbf{charin()} BIM can correctly interpret the character count for Unicode input.

Handling seek for variable-width encodings and EGCs is problematical.
\cref{issue:RA}
There are several viable approaches, each with disadvantages.

\begin{itemize}
\item Use octet count, as at present. \\
This risks a program doing a seek within a UTF-8 sequence,
within a UCS-2/UCS-4 character,
within a UTF-16 surrogate pair,
or, for EGC access, withing an EGC.
Strategies to ameliorate this include
\begin{definition}
\item [UTF-8]
Read the octet at the specified offset and verify that either the high
bit is 0 or the two high order bits are 1.
\item [UCS-2]
Verify that the offset is a multiple of 2.
\item [UTF-16]
Verify that the offset is a multiple of 2.  Read two octets at the
specified offset and verify that is not a low surrogate (U+DC00-U+DFFF),
\item [UCS-4]
Verify that the offset is a multiple of 2.
Read four bytes and verify that they do not exceed U+10FFFF.
\end{definition}
\item Use character or EGC count \\
This is expensive
\item Define a cursor type. \\
This violates "Everything is a string" in classic Rexx.
\end{itemize}

Handling update in the middle of a file with variable-width encodings
or EGC access has similar issues, Absent an acceptable way to deal
with it, prohibit update in the middle of a file for anything but
binary and legacy characters.

@@

Case folding of symbols and caseless comparisons will use the \cite{UCD}.
\cref{issue:CF,issue:CComp,issue:CConv})
The rules for case folding should take into account mathematical usage.
\begin{itemize}
\item U+1D400 through U+1D7FF are semantically distinct from other letters.
\item Superscripts and subscripts have semantic significance.
\end{itemize}
